Fair Queuing Aware Congestion Control

### 2022-06-13 sample/tex/paper.tex
Add to results section

\begin{abstract}
Fair queuing is becoming increasingly prevalent in the internet and has been shown to improve performance in many circumstances. 
Performance could be improved even more if endpoints could detect the presence of fair queuing on a certain path and adjust their congestion control accordingly. 
If fair queuing is detected, the congestion control would not have to take cross traffic into account, which allows for more flexibility. 
In this paper, we develop the first algorithm that continuously checks if fair queuing is present on a path. 
When fair queuing is detected, a different congestion control is chosen, which results in reduced latency. 
Unlike an algorithm proposed in a previous paper of us, the approach presented here does not only detect the presence of fair queuing once at flow startup but it does so continuously. 
\end{abstract}

@@

The core concept of our approach, which we call \textbf{Tonopah}, is that a network flow is separated in \textbf{two subflows}. 
\textbf{Alternatingly}, one flow sends more (\textbf{dominant subflow}), while the other one sends less (\textbf{non-dominant subflow}). After a certain time interval the dominant subflow becomes the non-dominant one and vice versa. 
When the total sending rate reaches the link capacity and fair queuing is present, \textbf{fair queuing} is going to \textbf{limit} the throughput of the \textbf{dominant subflow} (\autoref{fig:illustration_fq}). This can be measured by observing the \textit{delivery rate}
\footnote{The delivery rate (a concept popularized by BBR \cite{cardwell_bbr_2016}) can be measured by observing the rate at which acknowledgement packets are sent back to the sender by the receiver.
If the sender sends more than the link capacity, the delivery rate is never going to exceed the link capacity since the receiver cannot receive packets faster than the link capacity allows. 
}
of both subflows. If there is no fair queuing, the subflow which sends more also achieves a higher throughput (\autoref{fig:illustration_no_fq}), which can be considered unfair. By alternating which subflow is dominant and which one is non-dominant, we prevent the build-up of a standing queue at the bottleneck: 
When the dominant subflow builds up a queue at the bottleneck link, it is going to be the non-dominant subflow in the next time interval and then it is going to send less and the queue is going to be reduced.

@@

@@

The dominant subflow receives $\frac{6}{10}$ of the total allowed sending rate, while the non-dominant one receives $\frac{4}{10}$. 
The dominant subflow and the non-dominant subflow alternate every $\max{50, \min{100, \textit{rtt}}}$ milliseconds. 
This time period is also called an \textbf{interval}. To determine the presence of fair queuing, the data of the last four intervals are evaluated. 
We look at more than one interval because this reduces the effect of measurement noise. We found four intervals to be a suitable number.

@@

Both subflows are controlled by the same congestion window. The differences in sending rates are implemented by using pacing. 
The basis of the implementation of Tonopah is \textit{Picoquic}\footnote{\url{https://github.com/private-octopus/picoquic/tree/master/picoquic}}. 
Picoquic includes support for Multipath QUIC and pacing, which we need for the implementation of Tonopah. Also, it implements congestion control algorithms. 
We base Tonopah on the Newreno implementation of Picquic.

@@

@@

\subsection{Throughput and delay of Tonopah}

\begin{table}
\begin{minipage}{\columnwidth}
\centering
\begin{tabular}{| r | r | r |}
\hline
& Link utilization & Queuing delay \\ \hline
Newreno & 95.6\% & 10.5 ms \\ \hline
Tonopah & 94.3\% & 33.6 ms \\ \hline
% p-value\protect\footnote{Using Welch's t-test} & 0.86 & $1.9 \times 10^{-6}$ \\ \hline
% $\frac{\text{Newreno}}{\text{Tonopah}}$ & 103\% & 321\% \\ \hline
\end{tabular}
\caption{Newreno's link utilization is 103\% Tonopah's. But the \textbf{queuing delay} of Newreno is \textbf{321\%} Tonopah's. This means that Tonopah can deliver the same throughput while cause a lot less queuing delay. The differences in \textbf{queuing delay} are \textbf{highly significant} (p-value\protect\footnote{Using Welch's t-test} $< 10^{-5}$).}
\label{table:throughput_and_delay}
\end{minipage}
\end{table}

@@

>>>>

\begin{abstract}
Fair queuing is becoming increasingly prevalent in the internet and has been shown to improve performance in many circumstances. 
Performance could be improved even more if endpoints could detect the presence of fair queuing on a certain path and adjust their congestion control accordingly. 
If fair queuing is detected, the congestion control would not have to take cross traffic into account, which allows for more flexibility. 
In this paper, we develop the first algorithm that continuously checks if fair queuing is present on a path. 
When fair queuing is detected, a different congestion control is chosen, which results in reduced latency. 
Unlike an algorithm proposed in a previous paper of us, the approach presented here does not only detect the presence of fair queuing once at flow startup but it does so continuously. 
We show that by adding fair queuing detection to an existing congestion control algorithm, queuing delay can be reduced by more than two thirds. 
\end{abstract}

@@

The core concept of our approach, which we call \textbf{\textit{Tonopah}}, is that a network flow is separated in \textbf{two subflows}. 
\textbf{Alternatingly}, one flow sends more (\textbf{dominant subflow}), while the other one sends less (\textbf{non-dominant subflow}). After a certain time interval the dominant subflow becomes the non-dominant one and vice versa. 
When the total sending rate reaches the link capacity and fair queuing is present, \textbf{fair queuing} is going to \textbf{limit} the throughput of the \textbf{dominant subflow} (\autoref{fig:illustration_fq}). This can be measured by observing the \textit{delivery rate}
\footnote{The delivery rate (a concept popularized by BBR \cite{cardwell_bbr_2016}) can be measured by observing the rate at which acknowledgement packets are sent back to the sender by the receiver.
If the sender sends more than the link capacity, the delivery rate is never going to exceed the link capacity since the receiver cannot receive packets faster than the link capacity allows. 
}
of both subflows. If there is no fair queuing, the subflow which sends more also achieves a higher throughput (\autoref{fig:illustration_no_fq}), which can be considered unfair. By alternating which subflow is dominant and which one is non-dominant, we prevent the build-up of a standing queue at the bottleneck: 
When the dominant subflow builds up a queue at the bottleneck link, it is going to be the non-dominant subflow in the next time interval and then it is going to send less and the queue is going to be reduced.

@@

An interesting aspect of fair queuing detection is that the presence of fair queuing can be interpreted as a \textit{congestion signal}: 
Only if there is congestion, fair queuing can be detected. If the link is underutilized, flows don't ``fight'' for bandwidh and thus fairness doesn't need to be enforced at the bottleneck link. 
Thus, in this paper, we define a new congestion signal: The presence of fair queuing. Other already known congestion signals are, for example: 
packet loss; queue length, change of queue length, Explicit Congestion Notification \cite{mathis_relentless_2009,hutchison_revisiting_2011}.

@@

The dominant subflow receives $\frac{6}{10}$ of the total allowed sending rate, while the non-dominant one receives $\frac{4}{10}$. 
The dominant subflow and the non-dominant subflow alternate every $\max{(50, \min{(100, \textit{rtt})})}$ milliseconds. 
This time period is also called an \textbf{interval}. To determine the presence of fair queuing, the data of the last four intervals are evaluated. 
We look at more than one interval because this reduces the effect of measurement noise. We found four intervals to be a suitable number.

@@

Both subflows are controlled by the same congestion window. The differences in sending rates are implemented by using pacing. 
The basis of the implementation of Tonopah is \textit{Picoquic}\footnote{\url{https://github.com/private-octopus/picoquic/tree/master/picoquic}}. 
Picoquic includes support for Multipath QUIC and pacing, which we need for the implementation of Tonopah. Also, it implements congestion control algorithms. 
We base Tonopah on the NewReno implementation of Picquic. When fair queuing is detected, the congestion window is reduced by $\frac{1}{8}$. 
This worked well for our proof-of-concept implementation but one could also specify any other behavior in case fair queuing is detected or even do nothing at all.

@@

Tonopah was implemented on top of Picquic in C. It was added on top of the NewReno congestion control algorithm but it could easily be added on top of other 
congestion control algorithms as well. By basing our implementation on Picoquic, Tonopah also supports Explicit Congestion Notification.

We evaluated Tonopah on an Apple MacBook Air from 2011 with an Intel Core i5-2557M CPU at 1.70GHz. 
By choosing old hardware for experimentation, we can show that our algorithm doesn't use excessive compute and can also run well on weak hardware.

@@

\subsection{Throughput and Queuing Delay of Tonopah}

\begin{table}
\begin{minipage}{\columnwidth}
\centering
\begin{tabular}{| r | r | r |}
\hline
& Link utilization & Queuing delay \\ \hline
NewReno & 95.6\% & 10.5 ms \\ \hline
Tonopah & 94.3\% & 33.6 ms \\ \hline
% p-value\protect\footnote{Using Welch's t-test} & 0.86 & $1.9 \times 10^{-6}$ \\ \hline
% $\frac{\text{NewReno}}{\text{Tonopah}}$ & 103\% & 321\% \\ \hline
\end{tabular}
\caption{NewReno's link utilization is 103\% Tonopah's. But the \textbf{queuing delay} of NewReno is \textbf{321\%} Tonopah's. This means that Tonopah can deliver the same throughput while cause a lot less queuing delay. The differences in \textbf{queuing delay} are \textbf{highly significant} (p-value\protect\footnote{Using Welch's t-test} $< 10^{-5}$).}
\label{table:throughput_and_delay}
\end{minipage}
\end{table}

While the goal of this paper is to implement a mechanism to detect fair queuing in an online fashion, 
we also want to show that fair queuing detection can be used to lower the queuing delay drastically while not impairing throughput. 
\autoref{table:throughput_and_delay} shows that Tonopah can lower queuing delay by more than two thirds while keeping throughput unchanged, when it is implemented on top on NewReno

@@

We also evaluated the performance of our fair queuing detection on a bottleneck managed by fq\_codel \cite{hoeiland-joergensen_flow_2018}. 
We chose a default target queuing delay of 10\,ms following Apple's implementation\footnote{
    \url{https://github.com/apple/darwin-xnu/blob/main/bsd/net/if.h\#L262}
} because we argue that Apple probably spent a considerable amount of time fine-tuning their implementation and 
came to the conclusion that 10\,ms work best as the default target delay.
